---
title: "Simulating DNA copy number signals"
author: "P. Neuvial and D. Bontemps"
format:
  html:
    embed-resources: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
library("jointseg")
```

Generate data with known truth:

> for simplicity, only have two types of regions: one is centered around 0 and corresponds to 2 DNA copies (normal region), the other one corresponds to 3 DNA copies (region with a copy number alteration, CNA). The latter is not centered around 1 for biological/experimental reasons.

```{r signal}
## load known real copy number regions
affyDat <- acnr::loadCnRegionData(dataSet="GSE29172", tumorFraction=1)
dat <- subset(affyDat, region %in% c("(1,1)", "(1,2)"))
baseline <- mean(subset(dat, region == "(1,1)")$c) # mean of normal region

## generate a synthetic CN profile
K <- 10
n <- 1e4
sim <- getCopyNumberDataByResampling(n, K, regData=dat)
true_bkp <- sim$bkp # "bkp positions (the last row index before a breakpoint)"


Y <- sim$profile$c - baseline
plot(Y)
```

Segment (asking for twice as many segments as the true value)

```{r segmentation}
seg <- doDynamicProgramming(Y, K = 2*K)
seg$bkp
plotSeg(Y, breakpoints = seg$bkp)
```


Now we can build simultaneous confidence intervals for the mean of each segment, at a confidence level of .95

```{r conf-int}
library(posir)
delta <- max(10/n,0.005) # internal limitation of posir
posirCI <- posir::confidence_intervals_1D(Y,seg$bkp,delta,.05)
plot(posirCI)
```

Compute the empirical mean for each segment in the true segmentation, and add it to the previous plot

```{r segment-mean}
bx <- c(0, true_bkp, n) + 0.5
seg_means <- matrixStats::binMeans(Y, x = seq(along = Y), bx = bx)
plot(posirCI)
for(i in 1:K) abline(v=true_bkp[i]+0.5, col="blue")
for(i in 1:(K+1)) lines(c(bx[i],bx[i+1]), rep(seg_means[i],2), col="blue")
```

Comparison with a basic splitting technique : the observations in even position are used to select the segmentation, those in odd position to calculate the confidence intervals. If $k$ is the (potentially random) number of regions in the final segmentation, the confidence interval for the average value of $\mu$ in the region $I$ will have as radius $q_\hat{\sigma}/\sqrt{\ell(I)/2}$, where $\ell(I)$ is the length of $I$, the "/2" is due to the fact that only half of the observations are used, and $q$ is the Gaussian quantile of level $1-\alpha/(2k)$ (Bonferroni’s rule). We compare with the radius of POSIR confidence intervals $K_{1-\alpha,\delta} \hat{\sigma}/\sqrt{\ell(I)}$:

```{r comparison with Splitting+Bonferroni}
aux = function(k,p=0.95,delta=0.1) {
  return(sqrt(2)*qnorm(1-(1-p)/(2*k))/qposir(p,delta))
}
aux(1:10)
```

We see that for $\delta=0.1$ and a confidence rate of $95\%$, the POSIR method begins to be better as soon as the segmentation includes 7 cuts.

However, using the independence of observations, the different ICs are conditional independent to segmentation, and we can instead take the quantile of order $(1+\sqrt[k]{p})/2$:

```{r comparison with Splitting+conditional independence}
aux = function(k,p=0.95,delta=0.1) {
  return(sqrt(2)*qnorm((1+p^(1/k))/2)/qposir(p,delta))
}
aux(1:10)
```

Not much of a difference.

What if $\delta$ is chosen depending on the maximal number $K$ of intervals in the segmentation, for instance $\delta=\frac{1}{2 K}$?

```{r linking K and delta}
aux = function(k,p=0.95) {
  return(sqrt(2)*qnorm(1-(1-p)/(2*k))/qposir(p,1/(2*k)))
}
plot(1:100,aux(1:100),type="l")
```

```{r comparisons of quantiles for fixed K}
aux = function(c,k,p=0.95) {
  return(sqrt(2)*qnorm(1-(1-p)/(2*k))/qposir(p,c/k))
}
par(mfrow=c(2,2))
for(k in c(5,10,20,50)) {
  c=seq(from=k*.005, to=1, length.out=101)
  plot(c,aux(c,k),type="l", main=paste("Quantiles ratio for delta=c/K, with K =",toString(k)), 
       xlab = "c", ylab = "ratio")
}
```

## References

Pierre-Jean, M., Rigaill, G., & Neuvial, P. (2015). Performance evaluation of DNA copy number segmentation methods. Briefings in bioinformatics, 16(4), 600-615.
https://pubmed.ncbi.nlm.nih.gov/25202135/

Bontemps, D., Bachoc, F., & Neuvial, P. (2025). Inference post region selection. Preprint.
